---
title: "Kidney Disease PubMed"
author: "Janis Corona"
date: "12/11/2019"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

## This script takes articles from the abstracts on Kidney Disease articles from NCBI's PubMed, PLOS, and the summary of the NCBI GEO sample pages


This creates a directory to stem the abstracts and preprocess from the csv file
into a corpus of 20 files in a folder called KidneyDisease.
```{r, error=FALSE, message=FALSE, warning=FALSE}
Auto <- read.csv('NIH_PLOS_articles_kidney_disease.csv', sep=',',
                 header=FALSE, na.strings=c('',' '))
```

```{r, error=FALSE, message=FALSE, warning=FALSE}
colnames(Auto) <- c('abstract','source')
auto <- Auto[complete.cases(Auto$abstract),]


dir.create('./KidneyDisease')

ea <- as.character(auto$abstract)
setwd('./KidneyDisease')

for (j in 1:length(ea)){
  write(ea[j], paste(paste('EA',j, sep='.'), '.txt', sep=''))
}
setwd('../')


```

This code preprocesses and stems the corpus
```{r, error=FALSE, warning=FALSE, message=FALSE}
library(tm)
library(SnowballC)
library(wordcloud)
library(ggplot2)

KidneyDisease <- Corpus(DirSource("KidneyDisease"))


KidneyDisease

#KidneyDisease <- tm_map(KidneyDisease, removePunctuation)
#KidneyDisease <- tm_map(KidneyDisease, removeNumbers)
KidneyDisease <- tm_map(KidneyDisease, tolower)
KidneyDisease <- tm_map(KidneyDisease, removeWords, stopwords("english"))
KidneyDisease <- tm_map(KidneyDisease, stripWhitespace)
KidneyDisease <- tm_map(KidneyDisease, stemDocument)

dtmKidneyDisease <- DocumentTermMatrix(KidneyDisease)

freq <- colSums(as.matrix(dtmKidneyDisease))

```

This code orders words stemmed by frequency and finds input correlations
```{r}
FREQ <- data.frame(freq)
ord <- order(freq, decreasing=TRUE)

freq[head(ord, 25)]


```


```{r}
findAssocs(dtmKidneyDisease, "renal", corlimit=0.5)

```


```{r}
findAssocs(dtmKidneyDisease, "pain", corlimit=0.69)
```


```{r}
wf <- data.frame(word=names(freq), freq=freq)
p <- ggplot(subset(wf, freq>40), aes(word, freq))
p <- p + geom_bar(stat= 'identity') 
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1)) 
p

```


```{r}
wordcloud(names(freq), freq, min.freq=45,colors=brewer.pal(3,'Dark2'))

```


```{r}
wordcloud(names(freq), freq, max.words=30,colors=brewer.pal(6,'Dark2'))

```

### The above stemmed the corpus, this will lemmatize the original csv file
and add the field to the table and write out to csv, followed by plot the 
word count frequencies that were lemmatized and the word clouds

```{r, error=FALSE, message=FALSE, warning=FALSE}
library(textstem)

lemma <- lemmatize_strings(auto$abstract, dictionary=lexicon::hash_lemmas)

Lemma <- as.data.frame(lemma)
Lemma <- cbind(Lemma, auto)

colnames(Lemma) <- c('lemmatizedAbstract','abstract', 'source')

write.csv(Lemma, 'LemmatizedKidneyDisease.csv', row.names=FALSE)

```

```{r, error=FALSE, message=FALSE, warning=FALSE}
dir.create('./KidneyDisease-Lemma')

ea <- as.character(Lemma$lemmatizedAbstract)
setwd('./KidneyDisease-Lemma')

for (j in 1:length(ea)){
  write(ea[j], paste(paste('EAL',j, sep='.'), '.txt', sep=''))
}
setwd('../')

```

```{r, error=FALSE, message=FALSE, warning=FALSE}
library(tm)
library(SnowballC)
library(wordcloud)
library(ggplot2)

```



```{r}
KidneyDisease <- Corpus(DirSource("KidneyDisease-Lemma"))

KidneyDisease

#KidneyDisease <- tm_map(KidneyDisease, removePunctuation)
#KidneyDisease <- tm_map(KidneyDisease, removeNumbers)
KidneyDisease <- tm_map(KidneyDisease, tolower)
KidneyDisease <- tm_map(KidneyDisease, removeWords, stopwords("english"))
KidneyDisease <- tm_map(KidneyDisease, stripWhitespace)

dtmKidneyDisease <- DocumentTermMatrix(KidneyDisease)
dtmKidneyDisease

```

```{r}
freq <- colSums(as.matrix(dtmKidneyDisease))

FREQ <- data.frame(freq)
ord <- order(freq, decreasing=TRUE)

freq[head(ord, 25)]

```


```{r}
pain <- as.data.frame(findAssocs(dtmKidneyDisease, "pain", corlimit=0.99))

kidney <- as.data.frame(findAssocs(dtmKidneyDisease, "kidney", corlimit=0.65))


treatment <- as.data.frame(findAssocs(dtmKidneyDisease, "treatment", corlimit=0.81))

pain
kidney
treatment
```

```{r, width=500, height=500}
wf <- data.frame(word=names(freq), freq=freq)
p <- ggplot(subset(wf, freq>40), aes(word, freq))
p <- p + geom_bar(stat= 'identity') 
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1)) 
p

```

```{r}
wordcloud(names(freq), freq, min.freq=60,colors=brewer.pal(3,'Dark2'))

```

```{r}
wordcloud(names(freq), freq, max.words=30,colors=brewer.pal(6,'Dark2'))

```

